---
header-includes:
  - \usepackage{animate}
output: 
  pdf_document:
    keep_tex: true
fontsize: 12pt
geometry: margin=1in
number_sections: true
documentclass: article
---

# Appendix 9. Seasonal animation of predicted scoter abundance

To provide an example of the seasonal dynamics that can characterize sea duck abundance in Nantucket Sound, we produce an animation of predicted overall scoter abundance for every other day between 1 November 2005 and 31 March 2006.

```{r, echo=FALSE, eval=FALSE} 
# Generate the semi-daily data for the animation
source("../R/create_SCOT_animation_data.R")

# Compose the animation
source("../R/create_SCOT_animation.R")
```

The code associated with this .Rmd file produces `SCOT_2005_animation.pdf` which contains the final animation.  However, here we manually appended it to this Appendix by incorporating the following code into the associated .tex file and recompiling (see `Appendix_9__seasonal_dynamics_animation_.tex`)
```{r, echo=TRUE}
# \begin{figure}
# \begin{center}
# \animategraphics[controls, width=0.825\linewidth]{2}{Rplot}{}{}
# \end{center}
# \end{figure}
```