---
header-includes:
  - \usepackage{animate}
output: 
  pdf_document:
    keep_tex: true
fontsize: 12pt
geometry: margin=1in
number_sections: true
documentclass: article
---

# Appendix 9. Seasonal animation of predicted scoter occupancy and abundance

To provide an example of the seasonal dynamics that can characterize sea duck occupancy and abundance in Nantucket Sound, we proovide an animation of predicted scoter occupancy and average abundance for every other day between 1 November 2005 and 31 March 2006.

```{r, echo=FALSE, eval=FALSE} 
source("../R/utils.R")
source("../R/predict.gamlssHurdle.R")

instant_pkgs(c("plyr", "dplyr", "lubridate"))
load("../Data/ducks&environment.RData")

# Generate the semi-daily data for the animation
source("../R/create_SCOT_animation_data.R") # creates scot_anim

load("../Results_scot/zero.Rda")
scot_anim$occup <- mboost:::predict.mboost(zero, newdata=scot_anim, type = "response")

load("../Results_scot/hurdle.rda")
scot_anim$ltabund <- exp(predict.gamlssHurdle(zero, hurdle, scot_anim)$add_pred)

# Compose the animation
source("../R/create_SCOT_animation.R")
```

The code associated with this appendix (`Appendix_9_(seasonal_dynamics_animation).Rmd`) produces two files which contain the final animations: `SCOT_2005_occup_animation.pdf` and `SCOT_2005_abund_animation.pdf`.  However, here we manually appended them by incorporating the following code into the associated .tex file and recompiling (see `Appendix_9__seasonal_dynamics_animation_.tex`).
```{r, echo=TRUE}
# \begin{figure}
# \begin{center}
# \animategraphics[controls, width=1\linewidth]{2}{occup_anim}{}{}
# \end{center}
# \end{figure}

# \newpage

# \begin{figure}
# \begin{center}
# \animategraphics[controls, width=1\linewidth]{2}{abund_anim}{}{}
# \end{center}
# \end{figure}
```