---
output:
  pdf_document:
    includes:
      in_header: preamble.tex
      after_body: Figures.tex
#  word_document:  
fontsize: 12pt
documentclass: article
bibliography: NantucketMS.bib
csl: ecology.csl
---

```{r chunk-opts, echo=FALSE}
# Set some knitr chunk options
options(width = 250, scipen=999)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, echo=FALSE)
```


```{r load-data, include = FALSE}
source("../R/utils.R")

# Necessary packages; other packages may be loaded as other files are sourced
library(dplyr)

# Load sea duck data sets
# For each species/species group, load data sets for various analyses
# This creates the (1) final survey data set (e.g., coei), (2) zero-truncated
#   data set (i.e., at least 1 duck observed; e.g., coei_zerotrunc), and 
#   (3) presence/absence data set (e.g., coei_zero) 
source("../R/load_sea_duck_data.R")
```

RUNNING HEAD: Model-based distribution and abundance
\
\
\
Spatiotemporal modeling of winter sea duck abundance in Nantucket Sound, USA: implications for marine spatial planning 
\
\
<span style="font-variant: small-caps">
Adam D. Smith^1,^\footnotemark[6], Benjamin Hofner^2^, Jason E. Osenkowski^3^, Taber Allison^4^, Giancarlo Sadoti^5^, Scott R. McWilliams^1^, and Peter W. C. Paton^1^
</span>\
\
^1^Department of Natural Resources Science, University of Rhode Island, Kingston, Rhode Island 02881 USA

^2^Department of Medical Informatics, Biometry and Epidemiology, Friedrich-Alexander-University Erlangen-Nuremberg, Waldstraße 6, 91054 Erlangen, Germany

^3^Rhode Island Department of Environmental Management, 77 Great Neck Road, West Kingston, Rhode Island 02892 USA

^4^American Wind Wildlife Institute, 1110 Vermont Avenue, NW, Suite 950, Washington, D.C. 20005 USA

^5^Department of Geography, University of Nevada, Reno, Nevada, 89557 USA

\footnotetext[6]{Present address: United States Fish and Wildlife Service, National Wildlife Refuge System, Inventory and Monitoring Branch, 135 Phoenix Rd., Athens, GA 30605 USA; avianmigration@gmail.com}

\newpage

*Abstract*. Effective marine spatial plans require ecologically meaningful estimates of the distribution and abundance of biological resources that could be affected by anthropogenic change.  In North America, there is a pressing need to develop spatially-explicit abundance estimates of marine birds to help assess potential impacts of planned offshore wind energy developments (OWED). Sea ducks are especially important to consider in the context of marine spatial planning because populations of most North American species are below historic levels, and studies conducted in western Europe suggest that OWED could negatively impact sea duck populations.  Developers have proposed constructing 130 3.6-MW turbines in Nantucket Sound, Massachusetts where some of the largest concentrations of sea ducks occur during winter in eastern North America. We conducted aerial surveys from 2003 to 2005 to assess spatiotemporal variation in the abundance of sea ducks (Common Eider [*Somateria mollissima*], Black Scoter [*Melanitta americana*], Surf Scoter [*M. perspicillata*], White-winged Scoter [*M. deglandi*], and Long-tailed Duck [*Clangula hyemalis*]) in Nantucket Sound. We related spatiotemporal variation in sea duck occupancy and conditional abundance to potentially relevant biophysical and spatiotemporal covariates. We applied a negative binomial hurdle model that modelled occupancy using a generalized additive model (GAM), and generalized additive models for location, scale, and shape (GAMLSS) to estimate the conditional species abundance. Spatiotemporal effects (i.e., geographic location and time of season) were the dominant explanatory feature in occupancy and conditional abundance estimates.  Biophysical covariates also affected occupancy estimates, with occupancy greatest at intermediate monthly sea surface temperatures and in areas with coarser sediments; the effects of biophysical covariates on conditional abundance were less consistent among species. Resulting fine-scale, spatially- explicit models suggested that a planned OWED in Nantucket Sound could displace some sea ducks from potential foraging habitat; however, the largest concentrations of sea ducks were in areas away from the proposed OWED.  Our flexible, model-based approach can be used by local and regional planners to prioritize key habitats and areas used by marine birds and other fauna that should be protected from anthropogenic stressors such as OWED.  
\
\
*Key words*.  abundance, distribution, Common Eider, generalized additive models, gradient boosting, marine spatial planning, Long-tailed Duck, Nantucket Sound, scoter, sea ducks, wind energy 

\newpage

## Introduction

Marine spatial plans (MSP) can provide a rigorous framework to protect marine ecosystems from a variety of anthropogenic stressors including coastal development [@Polasky2008; @Foley2010].  A key element of MSPs is accurate spatially-explicit estimates of the abundance and distribution of biological resources, as well as information on the spatial distribution of potential human uses [@Douvere2008; @Punt2009; @Bradbury2014].  Along the northwest Atlantic Coast, there are numerous proposals to construct offshore wind energy developments [OWED; @Breton2009; @Musial2010], including the first in U.S. waters, the 5-turbine 30 MW facility in Rhode Island Sound to be operational by fall 2016. Therefore there is a pressing need to develop spatially-explicit models of key biological resources that can assist marine spatial planners in siting of OWEDs while minimizing impacts to marine ecosystem integrity.  

Understanding the distribution and abundance of wintering marine birds including sea ducks (Tribe: Mergini) represents a particular challenge to marine spatial planners because the environmental factors associated with their distribution and abundance remain notably understudied [@Zipkin2010; @Bowman2015; @Flanders2015]. Sea ducks are important to consider in the context of marine spatial planning because populations of most North American species are below historic levels and there is considerable uncertainty in population trajectories [@Bowman2015].  Moreover, evidence from Europe's numerous offshore wind facilities suggest that offshore wind energy developments could have negative consequences for sea duck populations [@Larsen2007; @Langston2013; @Bradbury2014].  

Model-based approaches are a useful framework to generate spatially-explicit estimates of animal abundance and changes in animal distribution independent of abundance [@Borchers2002; @Certain2008; @Nur2011; @Kinlan2012; @Winiarski2014]. However, modelling the spatial ecology of marine organisms such as sea ducks presents a variety of analytical challenges. First, the spatiotemporal distribution of marine organisms can be highly irregular and surveys often produce zero-inflated [@Martin2005; @Cunningham2005], overdispersed [@Richards2008] count data that may vary with biophysical features in complex, non-linear ways [@Austin2007]. Second, an adequate modeling approach usually must identify a small suite of important covariates among many potentially correlated covariates (i.e., high dimensionality; multicollinearity) while avoiding overfitting models, a process that often is complicated by data from repeated surveys that regularly exhibit spatial and temporal autocorrelation [@Hoeting2009].  

We demonstrate a model-based approach to predict animal distribution and abundance using aerial transect survey data of sea ducks in Nantucket Sound, Massachusetts, USA.  We implemented generalized additive models [GAMs; @Hastie1990; @Wood2006] and their recent extension [GAMLSS - GAMs for location, scale, and shape; @Rigby2005] in a gradient descent boosting framework [@Friedman2001]. GAMLSS methods extend GAMs to allow all parameters of the conditional response distribution to be modeled as a function of relevant covariates [@Rigby2005].  This approach provides spatially-explicit insights into the covariates associated with sea duck abundance and its variability. The boosting algorithm accommodates the inclusion of various effect types [e.g., linear, smooth, spatial, spatiotemporal, and random effects; @Hofner2014] and correlated covariates (i.e., parameter estimates are regularized), identifies the most relevant subset among potentially many covariates (i.e., variable selection), and evaluates competing representations of continuous covariates [e.g., linear vs. nonlinear effects of covariates; @Kneib2009; @Maloney2012]. Our specific objectives were to demonstrate a flexible, model-based approach that (1) provides spatially- and temporally-explicit estimates of sea duck abundance and distribution while accommodating many of the challenges inherent in animal survey data; and (2) describes the associations between biophysical features and sea duck distribution and abundance in one of the most important migrating and wintering areas for sea ducks in the western Atlantic.

## Methods

*Study area and species.*---We conducted fieldwork throughout Nantucket Sound in Massachusetts, USA (Figure 1). Our study area encompassed ca. 1,500 km^2^, was relatively shallow (generally < 20 m deep), and included some of the most important sea duck wintering habitat in the western Atlantic [@White2009; @Silverman2013].  The primary species of sea ducks found in Nantucket Sound were Common Eider (*Somateria  mollissima*; hereafter eider), Black Scoter (*M. americana*), Surf Scoter (*M. perspicillata*), and White-winged Scoter (*Melanitta deglandi*), and Long-tailed Duck (*Clangula hyemalis*).  Additionally, approximately 62 km^2^ of of Horseshoe Shoal in northwestern Nantucket Sound is fully permitted for OWED [Figure 1; @Santora2004], which could affect sea duck use of this important wintering area [@Drewitt2006; @Langston2013; @Winiarski2014].  

*Aerial strip transect surveys.*---During the winters (late October to mid-April) of 2003-2005, we conducted 30 (2003/2004: 13, 2004/2005: 10, 2005/2006: 7) standardized aerial strip-transect surveys [@Flanders2015] throughout an 1,100 km^2^ study area in Nantucket Sound (Figure 1).  Surveys occurred primarily from November - March (n = 27), with occasional October (n = 1) or April (n = 2) surveys. During each survey, we flew along 15 parallel (ca. 2.5 km apart), roughly north-south transects (Figure 1) using a high-wing, twin-engine aircraft (Cessna Skymaster 337) at an average altitude of 152 m and speed of 167 km/h (90 kts). This altitude allowed us to identify most birds at the sea surface and reduced sea disturbance (i.e., flushing birds to another part of the study area and, thus, potential double counting). The airspeed was the slowest at which the aircraft could safely fly. Surveys occurred only on days with wind speeds $\leq$ 15 kts and good visibility (>15 km). Surveys occurred between 0900 to 1600 h (average duration of ca. 2.5 h) to ensure that birds had completed their post-dawn movements [e.g., @Davis1997] but had yet to initiate pre-sunset movements from feeding to roosting areas; this time window also reduced glare due to low sun angles.

On each survey flight, two observers used their unaided eyes to continuously detect individuals or flocks, identifying sea ducks to species with the aid of binoculars as needed. Observers monitored the sea surface on their side of the plane in a ca. 91 m wide transect between ca. 56 - 147 m from the plane; observers could not see beneath the plane out to 56 m on each side. The narrow strip width ensured birds were detectable and identifiable with the naked eye and allowed us to avoid situations in which ducks were too abundant or spread over too wide an area to count accurately. This last point was especially relevant because sea ducks occasionally congregated in extremely large aggregations. We thus assumed inconsequential perception bias [i.e., few missed individuals present to be counted; @Marsh1989] given that our flight altitude and narrow transect width  [e.g., @Certain2008] within the strip transect [@Buckland2012].  Our transect dimensions resulted in the sampling of approximately 6% (median; 68.4 km^2^) of the study area during a survey. 

Observers verbally communicated duck sightings to a recorder, who entered the number of ducks, species, behavior (i.e., on the water or flying), and geographic location into a laptop computer linked to the plane's onboard GPS; we report results only from ducks observed sitting on the water. We subsequently consolidated counts for each species (eider and Long-tailed Duck) or species group (scoter) into 2.25km^2^ segments (Figure 1); this resolution (1.5 km x 1.5 km) corresponded approximately to the coarsest level of resolution of biophysical covariates (see below).

*Modeling approach.*---We related spatiotemporal variation in sea duck occupancy (i.e., probability of presence) and abundance to potentially relevant biophysical and spatiotemporal covariates. We applied a negative binomial hurdle model that first  modeled (1) the probability of occurrence of at least one individual (hereafter, the occupancy model) observed on the 180 m wide transect in a given segment using a logistic regression model and then (2) the abundance of sea ducks observed on the transect in that segment conditional on their presence (hereafter, the count model) using a truncated negative binomial model. The prevalence of zero counts (e.g., `r round(100 * nrow(coei[coei$count == 0,])/nrow(coei))`% of eider segment observations) prompted the use of a hurdle model. GAMs flexibly accommodated potential nonlinear effects of covariates on sea duck occupancy; GAMLSS accommodated potential nonlinear effects of covariates on the conditional mean and overdispersion of sea duck abundance in the count models.  We generated separate hurdle models for each sea duck group  [i.e., eider, scoter, and Long-tailed Duck; @Zipkin2010].

We implemented GAMs and GAMLSS in a gradient descent boosting framework. More specifically, we used component-wise functional gradient descent boosting to fit the models. We first computed the negative gradient of a user-specified loss function (typically the negative log-likelihood of the underlying model); the negative gradient of the loss function can be viewed as working residuals. We used the (negative) binomial log-likelihood as the loss function for occupancy models and the (negative) truncated negative binomial log-likelihood as the loss function for count models. 

In the next step, user-specified functional forms of the covariates relative to the response, called base-learners [@Hofner2014], were fitted separately to the negative gradient and a fraction of the single best fitting base-learner was added to the current model fit. The negative gradient was then reevaluated at the current model fit and the procedure iterated for a fixed number of iterations, *$m_\text{stop}$*, specified by the user [see @Buhlmann2007 for additional details]. We fit boosted GAMLSS models similarly, although in each iteration the negative gradient was computed separately with respect to each GAMLSS parameter (i.e., mean and overdispersion) while holding the other parameter as a fixed constant. We then fit base-learners to the resulting parameter-specific negative gradient vector and model updates were computed separately for each GAMLSS parameter as for boosted GAMs [see @Mayr2012 for additional details].

As only the single best-fitting base-learner was selected in each iteration, the algorithm thus integrated intrinsic selection of the most relevant covariates and their functional form (i.e., some base-learners may never be selected). Variable selection was further fostered by stopping the algorithm prior to convergence to maximum likelihood estimates [early stopping; @Mayr2012; @Maloney2012]. Furthermore, early stopping aims at maximizing predictive accuracy while avoiding model overfitting.  We used 25-fold subsampling to determine the optimal stopping iteration for each model. Specifically, we randomly drew (without replacement) 25 samples of size n/2 from the original data set. We used the selected sample to estimate the model and the balance of the data in each sample to determine the out-of-bag prediction accuracy (empirical risk) measured by the negative log-likelihood of each model; the optimal stopping iteration ($\hat{m}_\text{stop}$) is the iteration with the lowest average empirical risk. In boosted GAMLSS models we used multi-dimensional subsampling to determine the stopping iteration for each of the GAMLSS parameters while allowing for potentially different model complexities in the parameters; a detailed explanation of this cross-validation (subsampling) scheme is given in @Hofner2015b.

Despite these agreeable features, boosting methods typically produce "rich" models relying to some extent on many base-learners [@Hofner2015].  Thus, we additionally applied stability selection [@Meinshausen2010; @Shah2013] to identify those base-learners, and thus covariates, most commonly selected in each model while preserving an upper bound of $\alpha$ $\approx$ `r round(3/48, 2)` for the per-comparison error rate [see Appendix S1 and @Hofner2015 for more details].

*Covariates.*---We evaluated biophysical covariates expected to influence the distribution, abundance, and movements of sea ducks or, more likely, the distribution and availability of their benthic prey (e.g., mollusks and crustaceans; see Appendix S2); we did not have information related directly to the distribution of preferred prey.  Biophysical covariates could be characterized as strictly spatial or temporal effects (i.e., varying only among segments or within or among winters, respectively) or as spatiotemporal effects (i.e., varying among segments and within or among winters; see Appendix S2). Additionally, we included interactions that allowed the effects of two covariates to vary over time within a given winter (see Appendix S2).  We standardized (i.e., mean centered and scaled) all continuous covariates.  Appendix S3 describes a custom function to visualize the spatial and temporal distribution of covariates in Nantucket Sound.

For each continuous covariate, we specified two base-learners: a linear base-learner and a base-learner for the smooth deviation from the linear effect via penalized splines [i.e., P-splines; @Eilers1996; @Fahrmeir2004; @Schmid2008].  This allows the choice of the modeling alternative for each covariate, i.e., between no effect, linear effect, and smooth effect. We dummy-coded categorical covariates and used a separate linear base-learner for each of the *K* - 1 linear base-learners for a *K*-level covariate.  To address potential spatial autocorrelation, we included a smooth surface function of the spatial coordinates of segment centers [@Kneib2008]; this surface comprised four base-learners --- linear base-learners for the easting and northing, their linear interaction, and a penalized nonlinear tensor product P-spline [@Kneib2008; @Kneib2009; @Maloney2012].  We also allowed this surface to vary over time within a winter (via an interaction). @Hofner2014 describe base-learners in detail. 

The decomposition of continuous covariates into centered linear and penalized orthogonal nonlinear base-learners allowed us to weight base-learners equally and thus allowed unbiased model choice [i.e., prevented the preferential selection of smooth base-learners; @Kneib2009; @Hofner2011].  Specifically, we restricted each base-learner to a single degree of freedom and omitted the intercept term from each base-learner [@Kneib2009; @Hofner2011].  Consequently, we added an explicit linear base-learner for the model intercept. The occupancy and count models had the following structure (see Appendix S2 for descriptions of abbreviated covariates):

\begin{equation}
\begin{aligned}
g(\cdot) =  & int + time + f(time) + SSTw + f(SSTw) + SSTm + f(SSTm) + SSTrel +\\
     & f(SSTrel) + SSTrel \cdot time + f(SSTrel, time) + SBT + f(SBT) + NAOw + \\
     & depth + f(depth) + depth \cdot time + f(depth, time) + d2land + f(d2land) +\\
     & chla + f(chla) + cdom + f(cdom) + f(cdom, chla) + meanphi + f(meanphi) +\\
     & SAR + f(SAR) + tidebmean + f(tidebmean) + tidesd + f(tidesd) + strat +\\
     & f(strat) + ferry + y2004 + y2005 + xkm + ykm + xkm \cdot ykm + \\
     & f(xkm, ykm) + xkm \cdot time + ykm \cdot time + xkm \cdot ykm \cdot time + \\
     & f(xkm, ykm) \cdot time + obs\_window + f(obs\_window).  
\end{aligned}
\end{equation}

In our occupancy models, $g(\cdot)$ is $g(\pi_{\text{sea ducks}})$, the occupancy probability of a given duck species in a segment and $g$ is the logit link.  In our count models, $g(\cdot)$ took two forms within the GAMLSS framework -- the (conditional) mean count of sea ducks, $g(\mu_{\text{sea ducks}})$, and the (conditional) overdispersion in sea duck counts, $g(\sigma_{\text{sea ducks}})$; $g$ is the log link in both cases.  Base-learners denoted as $f(\cdot)$ indicate the penalized nonlinear deviations from the corresponding linear base-learner (e.g., $f(time)$). The explicit intercept (*int*) was a necessary byproduct of our decomposition of base-learners [see above and @Kneib2009; @Hofner2011].  Small values of *obs_window*, our measure of survey effort (see Appendix S2), in some segments impaired the estimability of models including *obs_window* as an offset; we thus included it as a covariate.

Subsequent to their independent fitting, we consolidated occupancy and conditional count models [see, e.g., Equation 6 in @Zeileis2008] to generate spatially-explicit estimates of unconditional sea duck abundance.  These estimates were used to evaluate the approximate explanatory power of our final models using a pseudo *R*^2^ measure of the explained variation [@Nagelkerke1991; @Maloney2012]. 

All analyses were conducted in R [Version 3.1.3; @RCoreTeam2014] with the add-on packages gamboostLSS [@Hofner2015a; @Hofner2015b], mboost [@Hothorn2010; @Hothorn2015], and stabs [@HofHoth2015].  The data and code for reproducing this manuscript and analyses are given as an online electronic supplement at [http://github.com/adamdsmith/NanSound_EcolApp](http://github.com/adamdsmith/NanSound_EcolApp).

## Results

We fit two models per each species because sea duck occupancy and conditional count were modelled independently. Bootstrapped empirical risk suggested that occupancy models for all species converged to the maximum likelihood estimates (i.e., occupancy models failed to stop early; see Appendix S4). Conversely, bootstrapped empirical risks prescribed early stopping for both the conditional mean and overdispersion parameter in all count models (see Appendix S4). Final occupancy models and models for the conditional mean of the count part included only a subset (12% to 38%) of the 48 base-learners initially specified for selection (see equation 1). Occupancy models generally contained more covariates and their interactions (8-10 of 23) than did count models (3-6 of 23), particularly among stably selected covariates and their interactions (Figure 2, see also Appendices S5 - S7).

*Sea duck occupancy.*---The covariates associated with occupancy were relatively consistent among sea duck species (Figure 2). The influence of univariate effects on the response is reflected in the range of the effect over the Y-axis and, due to standardization, can be compared among species and covariates within a model. For example, monthly sea surface temperature (*SSTm*) associated more strongly with eider occupancy than did distance to land (*d2land*) because it spanned a larger range of the Y-axis (Figure 2). In contrast, monthly sea surface temperature (*SSTm*) associated much more strongly with occupancy of Long-tailed Duck than eider and scoter for the same reason (Figure 2). Covariate interactions, illustrated with bivariate plots, are similarly comparable within a model. Only the general association (i.e., positive or negative) with the additive predictor is given for factor variables. Comparing univariate, bivariate, and categorical effects is accomplished using the detailed covariate plots for eider, scoters, and Long-tailed Duck (Appendices S5 - S7, respectively). 

Spatiotemporal effects (i.e., occupancy associated with the xkm-ykm location of segments and the change over time within winter [*time*]) were the dominant explanatory feature in occupancy models, although these patterns varied considerably among species (Figure 2; see Day of season, Northing x Easting). Occupancy increased, but at a decreasing rate, with survey effort (*obs_window*) in a given segment (Figure 2). Occupancy estimates increased at intermediate monthly sea surface temperature (*SSTm*), greater distances from land (*d2land*), and in areas with coarser sediments (i.e., smaller meanphi). Eider occupancy was associated negatively with chromomorphic dissolved organic material (*cdom*) and positively with sea floor surface area relative to planimetric area (*SAR*; our measure of the topographic variability of the sea floor; Figure 2), whereas scoter occupancy likewise related to *SAR* and *cdom*, but in the opposite direction in both cases (Figure 2). Scoter occupancy was modestly greater in deeper waters (*depth*), whereas Long-tailed Duck occupancy was greatest in shallow waters early in the winter but in deeper waters later in the winter (Figure 2; *depth* x *time* covariate). Other effects were relatively minor and inconsistent among species.

The strong association with occupancy of segment *xkm*-*ykm* location resulted in distinct spatial patterns of occupancy among species (Figure 3, top row) despite the relative similarity of occupancy associations with biophysical covariates (Figure 2). Occupancy was typically highest for eider in northwest and southwest Nantucket Sound, in interior Nantucket Sound for scoter, and in northeast and south Nantucket Sound for Long-tailed Duck (Figure 3, top row). All species tended to avoid the western edge of the Sound northeast of Martha's Vineyard. Generally, the areas of highest occupancy exhibited the lowest relative variability (Figure 3, bottom row), defined as the median absolute deviation (MAD) of occupancy relative to median occupancy within a segment (a measure analogous to the coefficient of variation).

*Sea duck conditional abundance and overdispersion.*---Spatial effects (*xkm*-*ykm*) were the dominant explanatory feature of conditional abundance estimates for  scoters and Long-tailed Duck, but they were not selected in the eider model (Figure 2). In contrast with the corresponding occupancy model, scoter conditional abundance decreased with increasing sediment grain size (*meanphi*).  Additionally, the relationships between eider conditional abundance and dissolved organic material (*cdom*) and sea floor topography (*SAR*; Figure 2) were more complex than with eider occupancy. The conditional abundance of eider and scoter was also associated with relatively warm or cool sea surface temperatures (*SST~rel~*; Figure 2).   Biophysical covariates associated with Long-tailed Duck conditional abundance exhibited general agreement with their counterpart in the occupancy models.

Spatially-explicit patterns of median conditional abundance (Figure 4, top row) did not necessarily reflect patterns of occupancy (cf. Figure 3, top row). Some areas of Nantucket Sound exhibited mutually high conditional abundance and occupancy for a given species (e.g., eider in the southwest, scoter in the interior, and Long-tailed Duck in parts of the northeast). However, conditional abundance was low despite relatively high occupancy in some instances (e.g., eider in the northeast and Horseshoe Shoal, scoter in the northeast and southeast, and Long-tailed Duck along the northern margin). Conversely, other areas of Nantucket Sound exhibited lower occupancy but sea ducks, when present, were more abundant (e.g., eider along the eastern margin, and scoter and Long-tailed Duck in the southwest). As in occupancy models, sea ducks were relatively absent from the middle-western margin of Nantucket Sound (i.e., northeast of Martha's Vineyard; see Figure 1). In contrast to sea duck occupancy, however, areas of highest conditional sea duck abundance typically exhibited the highest relative variability over time (Figure 4, bottom row).

Overdispersion in conditional sea duck abundance also varied with biophysical covariates, although there was less consistency in the associated covariates among species (Figure 2; see also Appendices S5 - S7). Variability (i.e., overdispersion) in sea duck counts was heterogeneous in space (Appendix S8; Figure S8.1, top row) and time (Appendix S8; Figure S8.1, bottom row) in Nantucket Sound, particularly for eider and scoter (as indicated by the magnitude of the overdispersion parameter values).  

```{r evaluate_models, eval=FALSE}
# How pseudo R2 was calculated
r2_packs <- c("pscl", "glmmADMB")
instant_pkgs(r2_packs); rm(r2_packs)
source("../R/pseudoR2.R")
source("../R/load_sea_duck_models.R") # Large files; can type a minute or two
(COEI_r2 <- round(pseudoR2(COEIocc, COEIcc, coei, engine = "pscl"), 2))
(SCOT_r2 <- round(pseudoR2(SCOTocc, SCOTcc, scot, engine = "pscl"), 2))
(LTDU_r2 <- round(pseudoR2(LTDUocc, LTDUcc, ltdu, engine = "pscl"), 2))
```

*Expected sea duck abundance.*---Consolidated occupancy and conditional count models provided estimates of unconditional sea duck abundance in the study area over the survey period (Figure 5, top row). Final models of expected sea duck abundance explained moderate amounts of variation in observed counts of eider, scoter, and Long-tailed Duck (pseudo *R*^2^ = 0.31, 0.48, and 0.32, respectively). Conditional abundance (Figure 4) strongly influenced the spatially-explicit patterns of expected abundance. Sea duck species exhibited relatively distinct patterns of abundance in Nantucket Sound. Eider were consistently most abundant in southwestern Nantucket Sound and also relatively abundant in the northeastern part of the study area but less consistently so as evidenced by the relatively high MAD/median abundance over time (Figure 5). Scoter were also most abundant, with occasional extremely large flocks, in southwestern Nantucket Sound, although this was also the area of highest relative variation in scoter abundance; relatively high abundances of scoter also occurred in interior Nantucket Sound (Figure 5). Long-tailed Ducks were consistently most abundant in northeastern Nantucket Sound, as well as along its southern margin (Figure 5). No species' highest abundances occurred in the permitted Nantucket Shoal area, although expected eider and scoter abundances were consistently elevated in some parts of the Shoal (west and southeast, respectively; Figure 5).

Summing the spatially explicit estimates of unconditional sea duck abundance (i.e., Figure 5) provides an estimate of total abundance in a 1.5 km x 180 m transect through all segments in the study area. We compared the total count (summed across all segments) of each sea duck species observed in aerial strip transects with the corresponding estimated total abundance in surveyed segments for each of the 30 aerial surveys (Figure 6). Our models tended to overestimate sea duck abundance when the actual numbers of sea ducks were relatively low, although overestimation was typically less than an order of magnitude. Additionally, scoter abundance was occasionally extreme relative to typical counts and somewhat prone to underestimation during these extreme counts. Nonetheless, the general adherence of observed and predicted abundance to a line of unit slope indicated that it may be reasonable to estimate sea duck abundance for the entire study area based on observed sea duck densities in transects (Figure 6).

*Temporal dynamics in wintering sea ducks.*---The MAD/median estimates (bottom rows of Figures 3-5, Figure S8.1 in Appendix S8) show that our spatially-explicit estimates of occupancy, abundance, and overdispersion invariably change over time, either explicitly via the selection of a within- or among-winter temporal effect (*time* and *y2004*/*y2005*, respectively) or implicitly via the selection of biophysical covariates that change within or among winters. The temporal dynamics of the wintering sea duck system in Nantucket Sound was one of its most striking attributes, and we illustrate these dynamics with an animation for scoter occupancy and abundance in Appendix S9. 

## Discussion

*Utility of a boosted GAMLSS modeling framework.*---We demonstrated a flexible model-based approach to evaluate the environmental associations of sea duck distribution and abundance based on multiyear replicated surveys. This approach is particularly useful for sea ducks that exhibit considerable variation in their spatial distribution and abundance within and between years [@Zipkin2010; @Winiarski2014; @Flanders2015]. The boosted GAMLSS framework offered several useful features including (1) the ability to model all parameters of the conditional distribution (e.g., conditional mean and overdispersion) as a function of covariates, (2) integrated variable reduction and selection among many covariates, and (3) integrated model selection via the simultaneous consideration of competing functional covariate forms (e.g., linear vs. non-linear). Additionally, this framework allowed us to incorporate smooth effects to efficiently account for spatiotemporal trends in the data that were poorly explained by other covariates and to identify those covariates and their functional forms most consistently associated with animal distribution and abundance (via stability selection). The useful features of this modeling framework apply especially to mobile species with non-uniform distributions that vary among and within years, such as the species of sea ducks that we studied.  Our estimates of the spatiotemporal abundance of sea ducks in Nantucket Sound were controlled largely by estimates of the conditional abundance and less by spatiotemporal patterns in the occupancy of sea ducks. This suggests that occupancy models alone may be inadequate for assessing risk from anthropogenic disturbances and for describing the fine-scale distribution of marine species [e.g., @Winiarski2014; @Flanders2015].

*The importance of spatial scale.*---The process whereby migratory animals such as sea ducks select a given area to inhabit during winter involves decisions at multiple spatial scales and the environmental attributes that determine this habitat selection often vary with spatial scale [@Johnson1980; @Johnson2004; @Johnson2006]. The majority of North American sea ducks migrate from high-latitude arctic and sub-arctic breeding areas to mid-latitude temperate wintering areas where they reside for most of the year [@Silverman2013; @Bowman2015; @Flanders2015]. At these large spatial scales, the distribution and abundance of sea ducks during winter may be affected by large-scale ocean characteristics [e.g., @Flint2013], climatic conditions [e.g., @Zipkin2010], and static or persistent habitat features [e.g., bathymetry, substrate, current and frontal systems; @Hyrenbach2000; @Nur2011; @Flanders2015].  At regional and local scales, however, most species of sea ducks congregate in large flocks (e.g., up to tens of thousands of birds) at sites where prey are abundant and accessible [@Loring2013; @Flint2013] although the abundance and distribution of these prey, and thus predators, can be extremely ephemeral and dynamic [@Hyrenbach2000; @Cisneros2011].

Given that sea duck distribution and abundance is spatially and temporally dynamic, yet expected to be driven by biophysical covariates [@Oppel2009; @Zipkin2010; @Flanders2015] that may differ in importance depending on spatial scale [@Johnson1980; @Johnson2004; @Johnson2006], marine spatial planners must carefully consider the most appropriate information to use when deciding, for example, where to place OWEDs. A larger-scale occupancy model developed by @Flanders2015 suggested that eiders were relatively uniformly distributed across Nantucket Sound, whereas our higher resolution abundance models found that eiders were concentrated in southwestern and central, eastern areas within Nantucket Sound.  While large-scale models [@Silverman2013; @Flanders2015] are useful to identify general geographic areas of import to sea ducks, our models provide more detailed estimates of sea duck distribution and abundance within Nantucket Sound. In terms of marine spatial planning, the large-scale models can be used to decide which areas should or should not be considered for OWED and, once an area is zoned for OWED, more localized models can be used to decide where best to place the OWED within the zoned area.

*Environmental covariates that best explain sea duck distribution and abundance.*---The biophysical associations with sea duck occupancy derived from our models were relatively consistent among species, whereas their associations with sea duck conditional abundance were more species-specific.  Distance to land, which often is positively associated with bathymetry, often has a strong influence on sea duck occupancy estimates [@Guillemette1993; @Lewis2008; @Winiarski2014; @Flanders2015].  Sediment grain size can also have a strong influence on prey availability for foraging sea ducks [@Goudie1988; @Lovvorn2009; @Loring2013] and affected occupancy and conditional abundance in this study.  In addition, topographic variability of the sea floor also influenced occupancy and conditional abundance, although the influence of topography on prey availability is less understood.  Sea surface temperature and chlorophyll *a* can have a strong influence on occupancy estimates for sea ducks [@Zipkin2010; @Flanders2015], although we found no effect of chlorophyll *a* during this study. Similarly, @Zipkin2010 found the North Atlantic Oscillation (NAO) was important in all these species of sea ducks at a continental scale, whereas we found no support that the NAO affected the distribution of sea ducks at the scale of this study. Certain covariates may associate with marine bird abundance or behavior at specific scales and not at others [@Logerwell1996] and this may explain the apparent discrepancy between studies in the effect of chlorophyll *a* and NAO. 

The unexplained variation in our models and the predominance of marginal spatiotemporal effects suggest that we probably missed important variable(s) relevant to the distribution of sea ducks in Nantucket Sound.  In addition, we likely need better biophysical proxies for the distribution of prey eaten by sea ducks or concurrent prey distribution information [e.g., @Vaitkus2001; @Kaiser2006; @Zydelis2009; @Cervencl2012; @Cervencl2014], although this is typically considerably more difficult to characterize at appropriate scales and does not guarantee improved predictive accuracy [@Gremillet2008; @Torres2008; @Benoit-Bird2013]. 

*Marine Spatial Planning: where to place a wind farm in Nantucket Sound given these estimates of the distribution and abundance of sea ducks?*---In the past decade, ecosystem-based marine spatial plans have become a reality because comprehensive land use planning is a central component of development plans in North America and Europe [@Douvere2008]. One of the biggest challenges facing marine spatial planners is the paucity of relevant information on the spatial distribution and abundance of biological resources including marine birds [@Bradbury2014; @Flanders2015].  We developed fine-scale, spatially- and temporally-explicit maps of the estimated distribution and abundance of sea ducks in Nantucket Sound that could assist marine spatial planners during the zoning process.  At a continental scale, Nantucket Sound regularly supports one of the largest concentrations of wintering sea ducks in eastern North America [@White2009; @Zipkin2010; @Silverman2013], therefore if any OWEDs are constructed in Nantucket Sound then steps should be taken to minimize impacts to this key wintering habitat.  @Bradbury2014 developed a sensitivity index that suggested that sea ducks were particularly vulnerable to habitat displacement from foraging sites, while they are less vulnerable to collision risk from OWED.  Our models suggest that the permitted OWED zone on Horseshoe Shoal is not located in prime foraging habitat for most species of sea ducks in the Sound, although large numbers of eiders can use this area [see also @Flanders2015].  Potential vulnerability to OWED has been incorporated into some recent spatially-explicit planning efforts for seabirds [@Garthe2004; @Bradbury2014; @Winiarski2014] though to our knowledge not yet for sea ducks. 

## Acknowledgments

We especially thank Simon Perkins, who had the primary role in coordinating field logistics, and John Ambroult, who piloted the majority of flights (and owned the plane).  Special thanks to Andrea Jones and Ellen Jedrey for their contributions to project design and field work.  Thanks also to Jack O'Brien and Phillip Kibler for piloting flights as needed, and to Jasmine Smith-Gillen and numerous volunteer observers and data recorders.  Funding was provided by Foundation M, The Island Foundation, Mass Audubon, Massachusetts Environmental Trust, the State of Rhode Island, the U.S. Fish and Wildlife Service Sea Duck Joint Venture, and the U.S. Fish and Wildlife Service Wildlife and Sport Fish Restoration Program.

## Literature cited
\hangparas{16pt}{1}