---
output: 
  pdf_document:
    toc: true
    toc_depth: 3
header-includes:
- \renewcommand{\contentsname}{S.3 Stable covariate effects}
#        includes:
#      in_header: preamble_app.tex
fontsize: 12pt
documentclass: article
---

```{r load_coei, echo=FALSE, message=FALSE, warning=FALSE}
source("../R/utils.R")
pacman::p_load(dplyr, ggplot2, lubridate)

## load plotting function
source("../R/full_effect_plot.R")

## Get covariate means and SDs
source("../R/create_new_data.R")

## Load sea duck data and relevant models
source("../R/load_sea_duck_data.R")
load("../Results_coei/zero.rda")
COEIocc <- zero
load("../Results_scot/zero.Rda")
SCOTocc <- zero
load("../Results_ltdu/zero.Rda")
LTDUocc <- zero

load("../Results_coei/hurdle.Rda")
COEIcc <- hurdle
load("../Results_scot/hurdle.Rda")
SCOTcc <- hurdle
load("../Results_ltdu/hurdle.Rda")
LTDUcc <- hurdle

## Set some graphic parameters
theme_set(theme_bw(base_size = 18))
theme_update(panel.grid.minor = element_blank(),
  panel.grid.major= element_blank(),
  panel.border=element_rect(fill=NA),
  panel.background= element_blank(),
  legend.position = 'none')

knitr::opts_chunk$set(message=FALSE, warning=FALSE, echo=FALSE,
                      fig.width = 6.5, fig.height=6.5)
```

&nbsp;
&nbsp;

Here we provide marginal functional plots of the relationships between stable covariates and the occupancy, conditional mean abundance, and conditional overdispersion of abundance of sea ducks in Nantucket Sound, Massachusetts, USA.  By 'marginal' we mean holding all other variables at their mean or reference values. Covariate plots are ordered roughly in descending order of the magnitude of their influence on the additive predictor in each model (or model parameter for count models).  Vertical lines along the *x*-axis (i.e., rug plots) indicate observed covariate values.  Covariates (and any abbreviations) are defined in detail in Table 1 in the manuscript; only effects selected to be stable are depicted (see Supplemental Information S1).

\newpage

## Common Eider

### Occupancy

```{r occ_coei, out.extra=sprintf('angle=%d', c(90, rep(0, 8)))}
# xkm, ykm
full_effect_plot(COEIocc, coei, var = "xkm, ykm", varType = "st", xlab = "Easting (km from center)",
                 ylab = "Northing (km from center")
cat('\r\n\r\n')
  
vars <- c("SSTm", "meanphi", "obs_window", "cdom", "SAR", "d2land", "SBT", "y2005")
xlabs <- c(expression(paste(SST, ", ", monthly~average~(degree*C))),
           "Sediment grain size (phi)", expression(Survey~effort~(transect~km^2)), 
           "Chromomorphic dissolved organic matter",
           "Sea floor surface area:planimetric area", "Distance to land (km)",
           expression(paste(SBT, ", ", May~-~Oct~(degree*C))), "Winter (2005)")
types <- c(rep("uni", 7), "fac")

for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(COEIocc, coei, var, type, xlab = xlab))
  cat('\r\n\r\n')
}
```

### Conditional mean
```{r mu_coei}
vars <- c("cdom", "SSTrel", "meanphi", "SAR", "depth", "ferry")
xlabs <- c("Chromomorphic dissolved organic matter", 
            expression(paste(SST, ", ", relative~to~other~segments~(degree*C))),
            "Sediment grain size (phi)", "Sea floor surface area:planimetric area", 
            "Depth relative to mean sea level (m)","Ferry route within 1 km")
types <- c(rep("uni", 5), "fac")
for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(COEIcc, parameter = "mu", var = var, varType = type, xlab = xlab))
  cat('\r\n\r\n')
}
```

### Conditional overdispersion
```{r sigma_coei, out.extra=sprintf('angle=%d', c(90, rep(0, 5)))}
full_effect_plot(COEIcc, parameter="sigma", var = "xkm, ykm", varType="bi", 
                 xlab = "Easting (km from center)",
                 ylab = "Northing (km from center)")
cat('\r\n\r\n')

vars <- c("tidebmean", "depth", "tidesd", "SSTm", "y2004")
xlabs <- c("Epibenthic tidal velocity (mean; m/s)", "Depth relative to mean sea level (m)",
            "Epibenthic tidal velocity (standard deviation; m/s)", 
            expression(paste(SST, ", ", monthly~average~(degree*C))), 
            "Winter (2004)")
types <- c(rep("uni", 4), "fac")
for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(COEIcc, parameter = "sigma", var = var, varType = type, xlab = xlab))
  cat('\r\n\r\n')
}
```

\newpage

## Black, Surf, and White-winged Scoter

### Occupancy

```{r occ_scot, out.extra=sprintf('angle=%d', c(90,rep(0, 10)))}
# xkm, ykm
full_effect_plot(SCOTocc, scot, "xkm, ykm", "st", xlab = "Easting (km from center)",
                 ylab = "Northing (km from center")
cat('\r\n\r\n')

vars <- c("obs_window", "time", "SAR", "SSTm", "cdom", "SSTw", "meanphi", "d2land", "depth", "SSTrel")
xlabs <- c(expression(Survey~effort~(transect~km^2)), "Days from 31 December", 
           "Sea floor surface area:planimetric area", 
           expression(paste(SST, ", ", monthly~average~(degree*C))), "Chromomorphic dissolved organic matter", 
           expression(paste(SST, ", ", Nov~-~Mar~(degree*C))),
           "Sediment grain size (phi)", "Distance to land (km)", 
           "Depth relative to mean sea level (m)",
           expression(paste(SST, ", ", relative~to~other~segments~(degree*C))))
types <- rep("uni", 10)

for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(SCOTocc, scot, var, type, xlab = xlab))
  cat('\r\n\r\n')
}
```

### Conditional mean
```{r mu_scot, out.extra=sprintf('angle=%d', c(90,0,0,0,0))}
# xkm, ykm
full_effect_plot(SCOTcc, parameter = "mu", var = "xkm, ykm", varType = "st", 
                 xlab = "Easting (km from center)", ylab = "Northing (km from center")
cat('\r\n\r\n')

vars <- c("SSTrel", "time", "d2land", "meanphi")
xlabs <- c(expression(paste(SST, ", ", relative~to~other~segments~(degree*C))),
            "Days from 31 December", "Distance to land (km)",
            "Sediment grain size (phi)")
types <- rep("uni", 4)
for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(SCOTcc, parameter = "mu", var = var, varType = type, xlab = xlab))
  cat('\r\n\r\n')
}
```

### Conditional overdispersion
```{r sigma_scot}
vars <- c("meanphi", "tidesd", "SAR", "SSTrel", "y2004", "y2005")
xlabs <- c("Sediment grain size (phi)", "Epibenthic tidal velocity (standard deviation; m/s)",
            "Sea floor surface area:planimetric area", 
            expression(paste(SST, ", ", relative~to~other~segments~(degree*C))),
            "Winter (2004)", "Winter (2005)")
types <- c(rep("uni", 4), "fac", "fac")
for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(SCOTcc, parameter = "sigma", var = var, varType = type, xlab = xlab))
  cat('\r\n\r\n')
}
```

\newpage

## Long-tailed Duck

### Occupancy
```{r occ_ltdu, out.extra=sprintf('angle=%d', c(90,0,0,0,0,90,0,0,0))}
# xkm, ykm
full_effect_plot(LTDUocc, ltdu, var = "xkm, ykm", varType = "st", xlab = "Easting (km from center)",
                 ylab = "Northing (km from center")
cat('\r\n\r\n')

vars <- c("SSTm", "time", "obs_window", "meanphi")
xlabs <- c(expression(paste(SST, ", ", monthly~average~(degree*C))),
          "Days from 31 December", expression(Survey~effort~(transect~km^2)),
          "Sediment grain size (phi)")
types <- rep("uni", 4)

for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(LTDUocc, ltdu, var, type, xlab = xlab))
  cat('\r\n\r\n')
}

full_effect_plot(LTDUocc, ltdu, var = "time, depth", varType = "bi",
                 xlab = "Days from 31 December", 
                 ylab = "Depth relative to mean sea level (m)")
cat('\r\n\r\n')

vars <- c("d2land", "tidesd", "y2004")
xlabs <- c("Distance to land (km)",
          "Epibenthic tidal velocity (standard deviation; m/s)",
          "Winter (2004)")
types <- c("uni", "uni", "fac")

for (i in 1:length(vars)) {
  var <- vars[i]
  xlab <- xlabs[i]
  type <- types[i]
  print(full_effect_plot(LTDUocc, ltdu, var, type, xlab = xlab))
  cat('\r\n\r\n')
}
```

### Conditional mean
```{r mu_ltdu, out.extra=sprintf('angle=%d', c(90,90,0))}
# xkm, ykm
full_effect_plot(LTDUcc, parameter = "mu", var = "xkm, ykm", varType = "st", 
                 xlab = "Easting (km from center)", ylab = "Northing (km from center")
cat('\r\n\r\n')

full_effect_plot(LTDUcc, parameter = "mu", var = "time, depth", varType = "bi", 
                 xlab = "Days from 31 December", 
                 ylab = "Depth relative to mean sea level (m)")
cat('\r\n\r\n')

full_effect_plot(LTDUcc, parameter = "mu", var = "SSTm", 
                 xlab = expression(paste(SST, ", ", monthly~average~(degree*C))))
cat('\r\n\r\n')
```

### Conditional overdispersion
```{r sigma_ltdu}
full_effect_plot(LTDUcc, parameter = "sigma", var = "strat", 
                 xlab = "Water column stratification potential (summer)")
cat('\r\n\r\n')
```